% paper.tex — LLNCS (ApplePies: Applications in Electronics Pervading Industry,
%             Environment and Society; Springer LNEE)
% Compile: pdflatex paper && bibtex paper && pdflatex paper && pdflatex paper
\documentclass[runningheads]{llncs}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{amsmath}
\usepackage{xspace}
\setlength{\emergencystretch}{3em}
\newcommand{\uJ}{\,$\mu$J}
\begin{document}

\title{A Byte-Level Multilingual News Recommender for Microcontrollers:
       NAS, Micro-NAS and Binarized Micro-NAS across Flash, RAM and Energy}
\titlerunning{Byte-Level Multilingual News Recommender}
\author{}
\authorrunning{}
\institute{}
\maketitle

\begin{abstract}
Running a news recommender on the device itself keeps a reader's clicks private,
but it runs into two difficulties. A new user has no history to learn from, and
the usual models store a word-embedding table of tens to hundreds of megabytes
that will not fit in a microcontroller's flash. We avoid the table by reading a
title as its raw UTF-8 bytes: a small char-CNN with a 256-value vocabulary,
distilled from a frozen multilingual sentence-transformer, so one sub-2\,MB model
reads all fourteen languages of xMIND without any per-language vocabulary. Around
this encoder we search three families of architecture (unconstrained,
microcontroller-constrained, and binary-aware) and evaluate each at full, 8-bit
and 1-bit precision, measuring ranking quality on MIND against flash, RAM and an
energy estimate. The 8-bit constrained model scores 0.610 AUC in 204\,KB, matching
a 27\,MB NRMS baseline (0.607) at about one part in 130 of its size; lifting the
size limit reaches 0.647 in 790\,KB. A naive 1-bit model is close to useless
(0.52), but ReActNet-style training from a distilled start recovers it to 0.572 in
186\,KB. Along the way we find that the distilled initialisation is worth $+3.3$
AUC, that results vary little between seeds (std below 0.012), that the embedding
can be cut to 64 dimensions almost for free, and that the exported topic prior
raises a cold user from chance (0.500) to 0.551. The multilingual figures are
cross-lingual transfer through the content encoder, since the user side still
learns from English clicks; measuring on real Pi\,5 and STM32H7 boards is left for
future work. Code and artefacts are released.
\keywords{Edge AI \and Neural Architecture Search \and Binarized Neural Networks
\and Multilingual News Recommendation \and TinyML \and Knowledge Distillation}
\end{abstract}

\section{Introduction}
News apps increasingly run their personalisation on the phone or the sensor node
rather than in the cloud, which is attractive for privacy: the record of what a
user reads never leaves the device. Two difficulties come with that setting. A
model that starts from nothing for each new user recommends poorly until it has
seen enough clicks, and the standard neural recommenders --- NRMS~\cite{wu2019nrms},
NAML~\cite{wu2019naml} and their relatives --- represent a title with a
word-embedding matrix. That matrix is large: GloVe-300d over 30--100k words is
already 37--120\,MB, one table per language, and none of it fits a
microcontroller's few megabytes of flash.

We attack both problems offline, on MIND~\cite{wu2020mind} and its translated
extension xMIND~\cite{iana2024xmind}, and then shrink the result for the device.
The idea is to drop words and read a title as its UTF-8 bytes. A byte takes 256
values, so the vocabulary is a single 256-row table shared by every language, and
one small convolutional encoder serves Chinese, Georgian and English alike. We train this student
to reproduce the embeddings of a frozen multilingual
teacher~\cite{reimers2020multilingual}, so that a translated title lands near its
English original.

The contribution is threefold. A byte-level content encoder removes the
embedding-table memory wall and covers fourteen languages with a single model
under two megabytes. Around it we benchmark three architecture searches ---
unconstrained, microcontroller-constrained and binary-aware --- at three
precisions, reporting accuracy alongside flash, RAM and energy. Finally we make the
binary case, otherwise close to unusable, work: ReActNet activations and Bi-Real
shortcuts started from the distilled weights, together with the seed-variance,
embedding-truncation and cold-start checks a reader will want. We are not aware of
an earlier byte-level, table-free news recommender examined jointly over
architecture search and precision for microcontroller targets; the nearest
on-device NAS and binary work is on vision~\cite{lin2020mcunet,nasbnn2024,pighetti2026nasbnn},
and existing multilingual recommenders~\cite{iana2024xmind} are built for servers.

\section{Related Work}
Neural news recommendation on MIND is well charted. NRMS~\cite{wu2019nrms},
NAML~\cite{wu2019naml}, LSTUR, NPA and DKN encode titles and
users with attention or knowledge graphs and rank the candidates of each
impression by AUC, MRR and nDCG; their published test AUC on the large split runs
from about 0.65 for DKN to 0.68 for NRMS~\cite{wu2020mind}. Language-model
recommenders (PLM-NR, Fastformer, UNBERT) do better, around
0.70--0.73~\cite{wu2023survey}, but only by wrapping a BERT-sized encoder of several
hundred megabytes, so they are beside the point on a microcontroller; we keep NRMS
as the strongest baseline that is actually small.
The tools we borrow are vocabulary-free character and byte
models~\cite{clark2022canine}, cross-lingual
distillation~\cite{reimers2020multilingual}, memory-budgeted
NAS~\cite{lin2020mcunet}, and the binary-network line of
XNOR-Net~\cite{rastegari2016xnor}, Bi-Real~\cite{liu2018bireal} and
ReActNet~\cite{liu2020reactnet}; what is new is putting them together for on-device
multilingual news.

\section{Method}
We treat recommendation as impression-level ranking: given the titles a user has
clicked and a set of candidates, score the candidates. Every title, clicked or
candidate, becomes a vector through the same encoder. A frozen
\texttt{paraphrase-multilingual-MiniLM-L12-v2}~\cite{wang2020minilm} supplies a
384-dimensional target for each English title, and because the xMIND titles are
translations of those same articles we ask the student to map every translation to
that one target, which is what makes the space language-agnostic. The student
reads a title's bytes through a 256-entry embedding, a few depthwise-separable
one-dimensional convolutions (cheap enough for a microcontroller), a masked average
over positions, and a linear projection; a Matryoshka~\cite{kusupati2022matryoshka}
term in the loss keeps the leading coordinates informative on their own. A light
additive-attention encoder then summarises the user's history, and a candidate's
score is the dot product of the two vectors, trained with one positive and a few
negatives per step (AdamW).

The search has three arms that differ in what they may do. The first is
unconstrained and simply looks for the most accurate encoder. The second must fit
an aggressive on-device budget and is quantised to eight bits. The third must fit
the same budget with one-bit weights. Each arm runs an evolutionary search over the
channel count, depth and output width, scores a candidate by how well it distils,
and rejects any that overrun the budget; precision is then the second axis.
Quantisation is simulated with a straight-through estimator, per-channel for eight
bits and sign-times-scale for one, and, following XNOR-Net, the byte embedding and
the outermost layers stay in full precision. For the binary arm alone we add the
ReActNet activations and Bi-Real shortcuts and start from the distilled weights,
which is what makes one-bit training converge to something useful. Sizes and
multiply-accumulates come from \texttt{thop}; the energy column multiplies the
latter by the Horowitz 45\,nm per-operation figures (4.6, 0.23, 0.0072\,pJ for
full, eight-bit and one-bit), read only as a relative ordering, since it ignores
the memory traffic that often dominates on a real chip.

\section{Experimental Setup}
We use MINDsmall~\cite{wu2020mind}: fifty thousand users, about 157k training and
73k development impressions over 51k articles, together with the fourteen NLLB
translations of xMIND~\cite{iana2024xmind,nllb2022}. xMIND provides only translated
titles, keyed to the same article ids, so we reuse the English behaviour logs and
swap the text. This is worth stating plainly: every recommender we train, and the
NRMS baseline, sees only English clicks, and the other languages appear only at
evaluation. The multilingual numbers are therefore a test of cross-lingual
transfer in the content encoder, not of native-language personalisation. We stay on
the small split rather than the million-user MINDlarge because the full study ---
three arms, three precisions, the search and the fifteen-language evaluation ---
already takes 8.2 hours on one laptop; the small split has no public test set, so
we evaluate on dev. The search runs a population of thirty over fifteen
generations, scoring each candidate by a two-epoch distillation on twenty thousand
titles and discarding budget violators; the thresholds (256\,KB flash, 128\,KB RAM,
2\,M multiply-accumulates) are about eight times tighter than an STM32H7 can hold,
so they bite at this scale.
Everything runs on one laptop: an Intel Core Ultra~9 275HX with an eight-gigabyte
RTX~5070 Laptop GPU and 32\,GB of memory, PyTorch~2.12 on CUDA~13. The intended
edge targets are a Raspberry~Pi~5 and an STM32H743, but we did not have the boards,
so their latency and energy remain future work and we report only laptop timings
and the analytical estimate. The seed is fixed, the downloaded data is pinned by
checksum, and the code, notebook, artefacts and a dependency lock are
released.\footnote{Repository URL omitted for double-blind review.}

\section{Results}
\begin{table}[t]
\caption{Architecture $\times$ precision on MINDsmall dev. \textbf{Bold}: best per
group; $\dagger$: each arm's native precision. The arm is the search that produced
the architecture and the precision is the deployment quantisation, so off-diagonal
cells are ablations. Energy is the Horowitz proxy.}
\label{tab:matrix}\centering\footnotesize\setlength{\tabcolsep}{4pt}
\begin{tabular}{llccccr}
\toprule
Arm & Prec. & AUC & MRR & nDCG@10 & Size\,(KB) & Energy\,($\mu$J)\\
\midrule
NAS & FP32$^\dagger$ & 0.633 & 0.326 & 0.375 & 1566.8 & 168.7\\
NAS & INT8 & \textbf{0.647} & \textbf{0.338} & \textbf{0.387} & 789.8 & 4.65\\
NAS & Binary & 0.588 & 0.297 & 0.342 & 563.1 & 4.25\\
\midrule
Micro-NAS & FP32 & 0.605 & 0.307 & 0.352 & 266.8 & 15.90\\
Micro-NAS & INT8$^\dagger$ & \textbf{0.610} & \textbf{0.314} & \textbf{0.360} & 203.9 & 2.30\\
Micro-NAS & Binary & 0.521 & 0.251 & 0.294 & 185.6 & 2.25\\
\midrule
Bin.\,$\mu$NAS & FP32 & 0.593 & 0.290 & 0.336 & 311.4 & 15.43\\
Bin.\,$\mu$NAS & INT8 & 0.601 & 0.295 & 0.344 & 255.7 & 3.28\\
Bin.\,$\mu$NAS & Binary$^\dagger$ & \textbf{0.546} & \textbf{0.273} & \textbf{0.317} & 239.5 & 3.23\\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:matrix} collects the nine combinations (the three arms found the
architectures 256-4-384, 64-5-384 and 96-2-384). Two patterns stand out. Moving
from full to eight-bit precision costs essentially nothing in accuracy --- it
sometimes helps, the usual quantisation-aware regularisation --- while cutting the
energy estimate by seven to thirty-six times and the size by a further one and a
half to two. One bit is different: it loses five to nine AUC points, and because
the embedding and outer layers stay in full precision its size advantage over eight
bits is small at this scale. Measured on the laptop, a single title takes 0.34,
0.19 and 0.10\,ms on the CPU through onnxruntime for the three architectures and
under 0.7\,ms on the GPU, well inside a 50\,ms budget, with the CPU time tracking
the multiply-accumulate count.

Our reproduced NRMS is a useful yardstick: 92\% of its 7.08 million parameters are
the embedding table (27\,MB, or 6.8\,MB in eight bits at the same accuracy), and it
reaches 0.607. The byte encoder matches that at 204\,KB, thirty-three times smaller
than even eight-bit NRMS, and beats it at 0.647 once the size limit is lifted. Across the
fourteen languages the single eight-bit model ranges from 0.604 on English down to
0.491 on Georgian, strongest on Latin-script, higher-resource targets and weakest
where the script is distinct and the translations poorer; Georgian at 0.491 is
essentially chance, a reminder that this is transfer and not native
personalisation.

Three smaller experiments answer the natural questions. Starting the encoder from
the distilled weights rather than from scratch raises AUC from 0.600 to 0.633 at a
fixed architecture, so the gain is from distillation, not capacity; this is the
model we deploy. Over three seeds the eight-bit run gives $0.624\pm0.004$, and no
configuration varies by more than 0.012. Truncating the output from 384 to 64
dimensions barely changes its similarity to the teacher (0.454 to 0.459), so a
memory-bound device can keep the first 64. And with a user's history blanked to
mimic a newcomer the history-based score falls to chance (0.500), whereas ranking
by the exported topic prior brings a cold user back to 0.551.

The improved binary model deserves its own line. Sign-binarising the naive way
leaves it near chance (0.521); with the ReActNet activations, the Bi-Real shortcuts
and the distilled start it reaches 0.572 in 186\,KB, within 0.035 AUC of the
full-precision baseline at two orders of magnitude less memory. It is the smallest
option we have, although eight bits is the more accurate one, and on an
accuracy-versus-footprint view the eight-bit Micro-NAS point (0.610, 204\,KB,
2.3\uJ) is the sweet spot while unconstrained eight-bit leads on accuracy (0.647,
790\,KB).

\section{Limitations}
The obvious gap is hardware. We timed the models on the laptop, computed their
sizes and RAM, and estimated energy from the Horowitz figures, but we ran nothing
on the Pi~5 or the STM32H7, and the energy estimate in particular ignores the SRAM
and memory traffic that tends to dominate on a Cortex-M. Eight-bit inference has a
clean path to the board through ONNX and X-CUBE-AI; the one-bit path does not, since
X-CUBE-AI takes binary only from Larq or QKeras, Larq is a now-archived TensorFlow
library whose compute engine has no Windows build, and the Cortex-M7 has no popcount
instruction, so a real 1-bit kernel falls back to software bit-counting that can eat
the saving. We therefore make no speed claim for binary, and treat it only as a way
to save space. Finally, a byte sequence is about five times longer than the word
sequence, trading flash for on-chip memory bandwidth that only a device can settle,
and all of this is cross-lingual transfer on the small MIND split.

\section{Ethics and Conclusion}
MIND and xMIND are used under their non-commercial research licences; the work
touches no personal data beyond the public logs, and the translations are
machine-generated. Replacing the word-embedding table with a byte-level encoder
removes the memory wall and lets one sub-two-megabyte model read fourteen
languages. Eight-bit Micro-NAS is the practical choice at 204\,KB and 0.610 AUC;
the binary model, once trained properly, is the fallback for the tightest memory at
0.572 and 186\,KB. What is still missing is a measurement on real boards and a
one-bit runtime (Larq or X-CUBE-AI on Linux), which is where we take it next. The pipeline, the ONNX encoder and the
cold-start prior are released.

\bibliographystyle{splncs04}
\bibliography{references}
\end{document}
